---
title: "Multinomial Regression for Correlated Data Using the Bootstrap in R"
author: Jennifer Thompson, MPH & Rameela Chandrasekhar, PhD, Vanderbilt University
date: "August 9, 2015"
fontsize: 13.5pt
output:
  beamer_presentation:
    incremental: false
---

```{r setopts, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = FALSE, error = FALSE, results='hide', fig.align='center', fig.pos='!h')

knitr::knit_hooks$set(mysize = function(before, options, envir) {
  if (before) 
    return(options$size)
})

options(width = 100)
```

# Purpose
  - Multinomial logistic regression: Useful for outcomes with >2 levels without inherent order
  - Model fits (# levels - 1) coefficients for each variable
  - Some methods exist in R, including:
    - VGAM package: `vglm()` with `family = multinomial()`
    - multgee package: `nomLORgee()`
  - To our knowledge, neither method allows us to easily get SEs/confidence intervals for predicted probabilities
    
# Proposed Method: Clustered Bootstrapped Multinomial Regression
  - Given data set with N subjects and $m_n$ records per subject, use clustered bootstrap sampling to create B data sets
    - Sample N subject IDs with replacement
    - Take all $m_n$ records from each sampled ID
  - Fit multinomial model on each of B data sets

# Proposed Method: Clustered Bootstrapped Multinomial Regression
  - *Coefficients:* Estimates = means of B estimates
  - *CIs:* Percentile method; ($2.5^{th}$, $97.5^{th}$)
  - *P-values:* Wald test
  - *Predicted probability* of an outcome level: Estimates straightforward; for CIs, use method in Liu's *Survival Analysis: Models and Applications* Appendix B
  - Functions collected in **[ClusterBootMultinom](http://github.com/jenniferthompson/ClusterBootMultinom/ "ClusterBootMultinom")** package on Github (github.com/jenniferthompson/ClusterBootMultinom)
  
```{r examplesetup, echo=FALSE}
## Load analysis data set
load('multibootstrap_data.Rdata')

## How many patients have >=1 day available?
n.pts <- length(unique(our.data$id))

library(ClusterBootMultinom)

# source('create_sampdata.R')
# source('multi_bootstrap.R')
# source('boot_coef_plot.R')

```

# Motivating Example
  - Cohort of critically ill patients with data collected daily in the ICU
  - Outcome: Mental status, assessed daily while in hospital; could be normal, delirious or comatose
  - Cannot assume that coma is worse than delirium
  - Exposure: Levels of a biomarker measured on study days 1, 3, and 5, if patient remained in the hospital
  - Most confounders also measured daily in the ICU
  - **Main question:** After adjusting for confounders, are biomarker levels associated with mental status on the day following biomarker measurement?
  - Final data: `r n.pts` unique patients with >=1 day of complete data; `r nrow(our.data)` total patient-days

```{r setuphide, echo=FALSE}
## Libraries used
library(VGAM)    ## for vglm()
library(ggplot2) ## for plotting
library(rms)     ## for using rcs() in model fits
library(aod)     ## for Wald tests
library(dplyr)
library(tidyr)   ## for faster data management
```

# Create Data Sets

```{r libraries, mysize=TRUE, size='\\small'}
# library(devtools)
# install_github('jenniferthompson/ClusterBootMultinom')

library(ClusterBootMultinom)
nboot <- 1000 ## Set number of bootstraps

```

Using **`create.sampdata()`**,

  - Create a list of B (here, `r nboot`) data sets, plus extra in case of nonconvergence
  - Each has all records from `r n.pts` IDs sampled with replacement from set of original IDs

```{r createsampdata, echo=TRUE, results='markup', mysize = TRUE, size='\\small'}
boot.datasets <-
  create.sampdata(org.data = our.data,
                  id.var = 'id',
                  n.sets = ceiling(nboot * 1.25))

```
```{r loadsampdata, echo=FALSE}
## Save to .Rdata file to save time on next run
save(boot.datasets,
     file = 'bootmulti_datasets.Rdata')

# load('bootmulti_datasets.Rdata')
```

# Run Models on Bootstrapped Data Sets

Using **`multi.bootstrap()`**:

  - Run model on original data set
  - If that model converges, run the same model on bootstrapped data sets until we reach B converged models
  - Save errors and warnings to .txt file
  - To calculate CIs for predicted probabilities for all outcome levels, run models twice, using highest & lowest outcome levels as reference

```{r multibootstrapform, echo=FALSE}
mod.formula.string <- 'mod.formula <- \n as.formula(mental.tmw ~ age + rcvd.ster +\n  sevsepsis + sofa.mod + as.factor(study.day) +\n  rcs(marker, 3))'

mod.formula <- as.formula(gsub('\\)$', '',
                               gsub('mod.formula <- \n as.formula(', '',
                                    gsub('\n  ', '', mod.formula.string, fixed = TRUE),
                                    fixed = TRUE)))

```
```{r printmultibootstrapform, echo=FALSE, results='markup', mysize=TRUE, size = '\\small'}
cat(mod.formula.string)
```

# Run Models on Bootstrapped Data Sets

```{r multibootstrapnormal, echo=TRUE, mysize=TRUE, size='\\small'}
## Run with Normal as reference level
boot.models.n <-
  multi.bootstrap(
    org.data = our.data,
      ## original data set
    data.sets = boot.datasets,
      ## list of bootstrapped data sets
    ref.outcome = grep('Normal', levels(our.data$mental.tmw)),
      ## outcome level to use as reference
    multi.form = mod.formula,
      ## model formula
    n.boot = nboot,
      ## number of successful model fits desired
    xvar = 'Marker')
      ## text for status updates

```
 **Returns** list: model fit on original data; list of `r nboot` successful model fits; number of times model failed to converge

```{r multibootstrapcoma, echo=FALSE}
# Run with Comatose as reference level
boot.models.c <-
  multi.bootstrap(org.data = our.data,
                  data.sets = boot.datasets,
                  ref.outcome = grep('Comatose', levels(our.data$mental.tmw)),
                  multi.form = mod.formula,
                  n.boot = nboot,
                  xvar = 'Marker')

## Get outcome comparisons for model sets to pass to plotting functions
get.out.comp <- function(org.mod){
  lapply(org.mod@misc$predictors.names, FUN = function(x){
    tmp <- strsplit(gsub('][\\)]*', '', gsub('[log\\(]*mu\\[,', '', x)), '/')
    return(paste(unlist(lapply(tmp, FUN = function(y){ levels(our.data$mental.tmw)[as.numeric(y)] })),
                 collapse = ' vs. '))
  })
}

out.comp.n <- get.out.comp(boot.models.n$org.model)
out.comp.c <- get.out.comp(boot.models.c$org.model)

```

```{r savebootcoefs, echo=FALSE}
## Keep only coefficients, as lists; create coefficient and vcov matrices
boot.coefs.n <-
  lapply(boot.models.n$boot.models,
         FUN = function(x){ return(x@coefficients) })
boot.matrix.n <- do.call('rbind', boot.coefs.n)
boot.vcov.n <- var(boot.matrix.n)

boot.coefs.c <-
  lapply(boot.models.c$boot.models,
         FUN = function(x){ return(x@coefficients) })
boot.matrix.c <- do.call('rbind', boot.coefs.c)
boot.vcov.c <- var(boot.matrix.c)

## Save original model fits
org.mod.n <- boot.models.n$org.model
org.mod.c <- boot.models.c$org.model

```
```{r loadbootcoefs, echo=FALSE}
save(org.mod.n, boot.matrix.n, boot.vcov.n,
     org.mod.c, boot.matrix.c, boot.vcov.c,
     file = 'boot_orgmodcoefs.Rdata')

# load('boot_orgmodcoefs.Rdata')
```

-----------

#### Check Distribution of Coefficients using **`boot.coef.plot()`** ####

```{r checkdistnormal, echo=FALSE, results='markup', include=TRUE, fig.height=4.25, fig.width=7, fig.align='center'}
## Histograms of bootstrapped coefficients
## Add reference lines for original and mean of bootstrapped coefficients

coefplot.n <- boot.coef.plot(coef.matrix = boot.matrix.n,
                             org.coefs = org.mod.n@coefficients,
                             plot.ints = FALSE)
coefplot.n$coef.plot + ggtitle('Reference = Normal')

```
```{r checkdistcoma, echo=FALSE}
# coefplot.c <- boot.coef.plot(coef.matrix = boot.matrix.c,
#                              org.coefs = org.mod.c@coefficients,
#                              plot.ints = TRUE)
# coefplot.c$coef.plot + ggtitle('Reference = Comatose')
```


# Calculate Odds Ratios

Use **`multi.plot.ors()`** to show ORs, 95% CIs for each outcome comparison.

```{r confoundersprep, echo=FALSE}
## Setup: Create data frame of variable labels to use in plots
or.labels <- data.frame(variable = c('age', 'rcvd.ster', 'sevsepsisSeverely septic today',
                                     'sofa.mod', 'as.factor(study.day)3',
                                     'as.factor(study.day)5'),
                        var.label = c('Age at enrollment',
                                      'Received steroid',
                                      'Severe sepsis',
                                      'Modified SOFA',
                                      'Study day 3 vs 1',
                                      'Study day 5 vs 1'))

# source('get_or_results.R')
# source('multi_plot_ors.R')

```

```{r confounderors, echo=TRUE, results='markup', mysize=TRUE, size='\\footnotesize'}
## Plot odds ratios and CIs for non-biomarker variables
covariate.ors <-
  multi.plot.ors(
    coef.list = list(boot.matrix.n, boot.matrix.c),
      ## List of matrices with bootstrapped coefs
    label.data = or.labels,
      ## data frame containing labels for each variable
    remove.vars = 'marker',
      ## this plot is just for confounders
    round.vars = 'age', round.digits = 3,
      ## round results for age to 3 instead of 2 places
    out.strings.list = list(out.comp.n, out.comp.c),
      ## list of strings describing comparisons
    delete.row = 'Normal vs. Comatose')
      ## One comparison will be redundant

```
**Returns** list: data frame with numeric results, ggplot2 object showing results

-----------
```{r confounderorplot, echo=TRUE, include=TRUE, fig.width=7, fig.height=5.5}
covariate.ors$or.plot
```


# Create Design Matrices

```{r designmats, echo=FALSE}
## Function to find mode of categorical variable
get.mode <- function(mode.var){ return(names(which.max(table(mode.var)))) }

## Create vector of adjustment values
adjvals <-
  c(1,                                                                    ## Intercept
    median(our.data$age, na.rm = TRUE),                                  ## Age
    as.numeric(get.mode(our.data$rcvd.ster)),                            ## Steroid use
    grep(get.mode(our.data$sevsepsis), levels(our.data$sevsepsis)) - 1, ## Severe sepsis
    median(our.data$sofa.mod, na.rm = TRUE),                             ## SOFA
    0, 0)                                                                 ## Study day = 1

## Create two rows, one for "outcome 1" and one for "outcome 2"
## Rows will later be repeated * number of unique biomarker values
design.out1 <- unlist(lapply(1:(length(adjvals)*2), FUN = function(i){
  if(i %% 2 == 1){ adjvals[i %/% 2 + 1] } else{ 0 } }))
design.out2 <- unlist(lapply(1:(length(adjvals)*2), FUN = function(i){
  if(i %% 2 == 0){ adjvals[i %/% 2] } else{ 0 } }))

```

To get predicted probabilities for outcomes vs. a continuous covariate, we need to adjust all other covariates to specific values.

  - Pass **`multi.plot.probs()`** [# outcome levels - 1] numeric vectors
  - Functions assume covariate in question is **last** variable in model formula; its X values will become columns at the end of design matrices
  - Example has $\beta_{0_{1, 2}}$ + 6 other $\beta$ per outcome level, excluding biomarker; set each to median/mode, representing "average" patient

```{r printdesignmats, echo=FALSE, results='markdown', mysize=TRUE, size='\\tiny'}
rbind(design.out1, design.out2)
```


# Calculate & Plot Predicted Probabilities, CIs of Each Mental Status by Marker Level

```{r predprobsprep, echo=FALSE}
# source('calc_spline.R')
# source('multi_calcppci.R')
# source('multi_plot_probs.R')

```

```{r predprobs, echo=TRUE, results='markup', mysize=TRUE, size='\\small'}
## Predicted probabilities for
##  outcome levels vs. biomarker
marker.prob.results <-
  multi.plot.probs(
    xval = 'marker',
    data.set = our.data,
    design.mat = list(design.out1, design.out2),
    mod.objs = list(boot.models.n$org.model,
                    boot.models.c$org.model),
    coef.list = list(boot.matrix.n,
                     boot.matrix.c),
    vcov.list = list(boot.vcov.n,
                     boot.vcov.c))

```
**Returns** list: data frame with numeric results, ggplot2 object showing results

----------
```{r printprobplot, echo=TRUE, include=TRUE, fig.height=4.75, fig.width=7}
marker.prob.results$prob.line.plot
```


# Simulation Study
  - Compared our method with
    1. `vglm()` from VGAM, without accounting for correlation
    2. `nomLORgee()` from multgee package, which accounts for correlation
  - Simulated 1000 data sets with correlated multinomial data, based on example from SimCorMultRes package
    - data sets included ID, time (cluster size = 3), one X ~ *N*(2.5, 3), outcome with *I* = 4 levels
    - all $\beta_{0i,...,I-1}$ = 1, all $\beta_{1i,...,I-1}$ = 2
    - correlation within patient = 0.9
    - N = 50, 150, 500

```{r exampledata, echo=FALSE, message=FALSE}
library(SimCorMultRes)

## Initialize number of patients, response categories, repeated measures per patient
npts <- 500
ncats <- 4
ntimes <- 3

## Initialize beta coefficients - keep it simple!
betas0 <- c(rep(1, ncats - 1), 0)
betas1 <- c(rep(2, ncats - 1), 0)

## Set correlation matrix
cormat <- toeplitz(c(1, rep(0, ncats - 1), rep(c(0.9, rep(0, (ncats - 1))), ntimes - 1)))

set.seed(1)
xmat <- matrix(rnorm(npts, 2.5, 3), npts, ncats)

lpmat <- matrix(betas1, npts, ncats, byrow = TRUE) * xmat + matrix(betas0, npts, ncats, byrow = TRUE)
lpmat <- matrix(lpmat, npts, ncats * ntimes)

yvals <- rmult.bcl(clsize = ntimes, ncategories = ncats, lin.pred = lpmat, cor.matrix = cormat)$Ysim

data <- data.frame(cbind(c(t(yvals)), c(t(xmat[,-ncats]))))
data$id <- rep(1:npts, each = ntimes)
data$time <- rep(1:ntimes, npts)
colnames(data) <- c("y","x","id","time")

```

# Model Convergence

Proportions of models which did not converge:

Method              | N = 50 | N = 150 | N = 500
------------------- |:------:|:-------:|:------:
`nomLORgee()`       |   0.49 |    0.18 |   0.08
`vglm()`            |   0.14 |    0.01 |   0.01
Clustered bootstrap |   0.33 |    0.03 |   0.01

# Relative Efficiency, Bias & CIs

![Combined results](simulated_allresults.pdf)

# Future Work & Acknowledgements
  - Future directions
    - Additional CI methods
    - Extending package to include more nonlinear terms, other flexibilities
  - Clinical investigators & coauthors:
    - Tim Girard, MD, MSCI
    - Pratik Pandharipande, MD, MSCI
    - Wes Ely, MD, MPH
  - R package resources:
    - Hilary Parker - [Writing an R Package from Scratch](http://hilaryparker.com/2014/04/29/writing-an-r-package-from-scratch/)
    - Hadley Wickham - devtools, roxygen2, *[R Packages](http://r-pkgs.had.co.nz/)*
    - Karl Broman - [R package primer](http://kbroman.org/pkg_primer/)
    - Jeremy Stephens, VUMC computer systems analyst
  - Email: jennifer.l.thompson@vanderbilt.edu
  - Package: [github.com/jenniferthompson/ClusterBootMultinom](http://www.github.com/jenniferthompson/ClusterBootMultinom)
